---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document: default
---

This is the a report that contains all the required information for "Peer Assessment 1" assignment.  
The raw data was downloaded from <https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip>.  
**The raw data has to be in the *work directory***  

The variables included in this dataset are:  
* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)  
* **date**: The date on which the measurement was taken in YYYY-MM-DD format  
* **interval**: Identifier for the 5-minute interval in which measurement was taken  

## Loading and preprocessing the data

First we will load the data.
```{r load data, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
unzip("activity.zip")
data <- read.csv("activity.csv")
```
  
Calculate number of total steps for each day and plot an histogram:
```{r calc total steps & plot histogram}
sumByDate <- summarise(group_by(data,date),TotalSteps = sum(steps,na.rm = TRUE))

hist(sumByDate$TotalSteps,breaks = 50, 
     main = paste("Histogram of total steps in day"), 
     xlab = "Total Steps")
abline(v=mean(sumByDate$TotalSteps),col="blue",lwd=2)
abline(v=median(sumByDate$TotalSteps),col="green",lwd=2)
legend("topright",
        c("mean","median"),
        lty=c(1,1),
        lwd=c(2.5,2.5),col=c("blue","green"))
```
  
The mean and median of the total number of steps were calculated as following:
```{r calc mean and median}
mean(sumByDate$TotalSteps)
median(sumByDate$TotalSteps)
```
  
## Imputing missing values

In this part we were required to plot an average of steps taken as function of 5-minute intervals.
Therefore, we will first group the data by 5-minute invervals and then calculate an average per interval.
```{r group by intervals and calc average}
averageByInterval <- summarise(group_by(data,interval),average = mean(steps,na.rm = TRUE))
```
  
The 5-minute interval in which the average of total steps is maximal is:
```{r max interval calc}
averageByInterval$interval[which.max(averageByInterval$average)]
```
  
Using the calculated data we will plot the average number of steps per 5-minute interval
```{r plot average per interval}
plot(averageByInterval$interval,averageByInterval$average,type="l",
     main = "Graph of average steps per interval",
     xlab = "5-minute interval", ylab = "Average number of steps")
```
  
As requested, number of rows containing "NA" is given by:
```{r calc number of NAs}
sum(is.na(data$steps))
```
  
I will replace the number of steps where the steps is "NA" in an average value of steps for the related interval of the replaced row.  
**dataImputed** is the variable name that will hold the same data holded in original table **data** but with imputed missing values
```{r Imputind missing values}
dataImputed <- data
naIndices <- which(is.na(data$steps))
for (i in naIndices) {
    dataImputed[i,"steps"] = averageByInterval[which(averageByInterval$interval == data[i,"interval"]),"average"]
}
```
  
Now let's repeat the plots and calculations on the new data (with imputed missing values):
```{r plot histogram for new data}
sumByDateImputed <- summarise(group_by(dataImputed,date),TotalSteps = sum(steps,na.rm = TRUE))

hist(sumByDateImputed$TotalSteps,breaks = 50, 
     main = paste("Histogram of total steps in day - With imputed missing values"), 
     xlab = "Total Steps")
abline(v=mean(sumByDateImputed$TotalSteps),col="blue",lwd=2)
abline(v=median(sumByDateImputed$TotalSteps),col="green",lwd=2)
legend("topright",
        c("mean","median"),
        lty=c(1,1),
        lwd=c(2.5,2.5),col=c("blue","green"))
```
  
The mean and median on the data with the imputed missing values were calculated as following:
```{r calc mean and median imputed}
mean(sumByDateImputed$TotalSteps)
median(sumByDateImputed$TotalSteps)
```
  
  
Results comparison:

data type|   mean    |   median
---------|-----------|----------
original |   `r mean(sumByDate$TotalSteps)`     |  `r median(sumByDate$TotalSteps)`
imputed  |   `r mean(sumByDateImputed$TotalSteps)`     |  `r median(sumByDateImputed$TotalSteps)`

According to the results in the comparison table presented above we can see that imputation of the missing values caused two modifications:  
1. The mean value was changed  
2. The mean and median values got equal (due to the type of imputing that was picked)
  
## Presentation as function of day type

I will create a new variable that is a replicate of the data with imputed missing values.  
Then add a new column called *day_type* that will be *weekend* or *weekday* according to the date.

```{r data with weekdays}
dataWeekdays <- data
dataWeekdays[,"day_type"] <- rep(0,nrow(dataWeekdays))
dataWeekdays[,"day_type"] <- (weekdays(as.Date(dataWeekdays$date)) == "Saturday") |
                             (weekdays(as.Date(dataWeekdays$date)) == "Sunday")
dataWeekdays[dataWeekdays$day_type == FALSE,"day_type"] <- "weekday"
dataWeekdays[dataWeekdays$day_type == TRUE,"day_type"] <- "weekend"

```
  
At last, I will group the new constructed data by weekday/weekend type and then calculate mean for each of 5-minute intervals:
```{r day type plot with panels}
dataWeekdaysGrouped <- summarise(group_by(dataWeekdays,day_type,interval),average = mean(steps,na.rm = TRUE))
    
qplot(interval,average,data=dataWeekdaysGrouped,geom="line",facets=.~day_type,
      ylab="average number of steps", xlab="5-minute invterval")
```
